---
title: "Load and Normalize the qPCR Output"
author: "Otho Mantegazza"
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
    toc: yes
    number_sections: true
    highlight: default
vignette: >
  %\VignetteIndexEntry{Introduction to Fluidgr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This package will help you to analyze the output of the [Fluidigm digital real time qPCR machine](https://www.fluidigm.com/reagents/digital-pcr) to inspect gene expression. You can do this with a simple workflow, that will be explained in detail further down this document.

# Setup

```{r setup, message = FALSE}
library(fluidgr)
library(magrittr)
library(dplyr)
library(stringr)
```

# Simple Workflow

Record the path to your fluidigm CSV data and the name of your reference normalizers.

```{r}
path_to_data <- system.file("extdata", "sample-fluidigm-run.csv",
                            package = "fluidgr",
                            mustWork = TRUE)

normalizers <- c("normalizer1",
                 "normalizer2",
                 "normalizer3")
```


Parse them, normalize them and scale them in a [magrittr](https://magrittr.tidyverse.org/index.html) pipe.

```{r, eval=FALSE}
fluidigm_data <- 
  path_to_data %>%
  read_fluidigm() %>%
  normalize() %>%
  scale_fluidigm()
```

And your data are ready to be saved on disk or plotted with ggplot2.

In the next paragraphs you'll find this same workflow explained in details.

# Load your data into R

You can start by loading the output of the Fluidigm digital real time qPCR machine into R.
This output is in a standard CSV format with a 10 lines header and triple column names, you can find [an example of such file here](https://github.com/othomantegazza/fluidgr/blob/master/inst/extdata/sample-fluidigm-run.csv).

Access the path to that sample CSV file in R with:

```{r}
path_to_data <- system.file("extdata", "sample-fluidigm-run.csv",
                            package = "fluidgr",
                            mustWork = TRUE)
```

And load it with the funciton `read_fluidigm()`. Use `simplify = TRUE` if you want to keep only the columns that are strictly necessary.

```{r}
dat <- read_fluidigm(path = path_to_data,
                     simplify = TRUE)
```

## Tidy your Data

After you have parsed the data, you can modify them as you wish using [dplyr](https://dplyr.tidyverse.org/articles/dplyr.html).

### Remove samples

For example, you might want to remove selected samples:

Our example dataset contains samples that are dilution used for calibration curves.

```{r}
dat$sample_name %>% unique()
```

Those samples, all have the word "Mix" in their name, and can be removed with:

```{r}
dat <- 
  dat %>%
  filter(!sample_name %>% str_detect("Mix"))
```

One sample "H2O" is a negative control, it also can be removed:

```{r}
dat <- 
  dat %>%
  filter(sample_name != "H20")
```

### Check measurements per sample

You can check that each sample has been measured the same number of times:

```{r}
dat$sample_name %>% table()
dat$sample_name %>% table() %>% unique()
```

### Check for duplicated targets

Also, one of the target is a duplicated normalizer. You can remove it with:

```{r}
dat$target_name %>% unique()

dat <- 
  dat %>%
  filter(target_name != "normalizer1bis")
```

## Load data manually

You can load the digital qPCR gene expression data manually with your favourite functions, and store them as a `tibble` or a `data.frame` in R.

For the analysis downstream, three columns are essential:

- `sample_name`, where you must store a univocal sample identifier.
- `target_name`, where you must store a univocal target identifier (i.e. the gene that you amplify).
- `ct_value`, where you must store the Ct value (i.e. threeshold cycle at which the amplicon is detected) as provided by the digital qPCR machine.

# Normalize

You must provide the names of the normalizers exactly as they are stored in the `target_name` column. In this case the name of the normalizers is: `normalizer1`, `normalizer2` and `normalizer3`.

```{r}
normalizers <- c("normalizer1",
                 "normalizer2",
                 "normalizer3")

norm_dat <- 
  dat %>% 
  normalize(normalizers = normalizers) 
```

The `normalize()` function takes the data parsed by `read_fluidigm()` as object and returns the same data with two additional columns:

- `norm_geom_mean` stores the geometric mean of normalizers for each sample.
- `expression` stores the normalized expression values.

Expression values are normalized with the formula: 

$$expression = 2^{-(C_T - C_Tnorm)}$$

Where, for each well, $C_T$ is the recorded threshold cycle (`ct_value`) and $C_Tnorm$ is the geometric mean of the normalizers C~T~ values.

This formula is a simplification of the common $2^{ - \Delta \Delta Ct}$ formula that skips calibration.

This is how your data should look after normalization:

```{r}
norm_dat %>% head() %>% knitr::kable()
```


# Session Info

```{r}
devtools::session_info()
```
