---
title: "Delta Delta Ct, Center on One Experimental Condition"
author: "Otho Mantegazza"
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
    toc: yes
    number_sections: true
    highlight: default
vignette: >
  %\VignetteIndexEntry{Delta Delta Ct, Center on One Experimental Condition}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Here, you will learn how to scale, inspect and visualize the gene expression data by centering the mesurement on one single experimental condition (in what is known as the $2^{ - \Delta \Delta Ct}$ method).

This method is uesful when you want to take one experimental condition as reference. For example, the start time when analyzing time series.

# Setup

```{r packages, message = FALSE}
library(fluidgr)
library(magrittr)
library(dplyr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(devtools)
```

# Data

We use the `scaled_fluidigm` dataset to illustrate how to apply the $2^{ - \Delta \Delta Ct}$ method.

```{r}
data(scaled_fluidigm)
```

# How to use the function `scale_ddct()`

```{r}
scaled_fluidigm %>% 
  scale_ddct(group_var = species,
             group_var2 = target_name,
             compare_var = stage,
             center_on = "N1")
```

