---
title: "Scale and Visualize qPCR results"
author: "Otho Mantegazza"
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
    toc: yes
    number_sections: true
    highlight: default
vignette: >
  %\VignetteIndexEntry{Scale and Visualize qPCR results}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, you will learn how to scale, inspect and visualize the gene expression data that you have parsed in the previous vignette (Load and Normalize the qPCR Output).

# Setup


```{r packages, message = FALSE}
library(fluidgr)
library(magrittr)
library(dplyr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(devtools)
```

# Data

We provide the `fluidigm_expr` object, that contains normalized gene expression measurement. 

This object is exactly the same at the one produced in the vignette (Load and Normalize the qPCR Output).

You can load it with:

```{r}
fluidigm_expr
```

# Scale

When you want want to explore gene expression, scaling is not always necessary, and often you can use, explore, visualize and communicate directly the normalized expression values; stored in the `expression` column of the `fluidigm_expr` dataset.

Indeed scaling is not always necessary, but, together with exploratory plots, scaling can help you detect patterns in your data.

Exploratory plots can help understand the effect that scaling has on expression data. This is why we explain scaling and visualisation of expression data together.

### Scale with the `scale_fluidigm()` function

Fluidgr provides the `scale_fluidigm()` function, that scales expression values by z-score according to the groups provided in the argument `.group`.

This function takes as input the output of `normalize()` and returns the same dataset with the extra column `scaled_expression`.

```{r}
scaled_dat <- 
  fluidigm_expr %>% 
  scale_fluidigm(.group = target_name)
```


### Write your own scaling function

You can write your own scaling function, becaus `scale_fluidigm()` function is just a soft wrapper around dplyr's `group_by()` and `mutate()`. If you want to try different scaling methods, you can replicate and tweak the output of `scale_fluidigm()` with:

```{r, eval = FALSE}
fluidigm_expr %>%
  group_by(target_name) %>%
  mutate(scaled_expression = scale(expression))
```

This should provide to you a good basis and enough freedom to explore and visualize your data as you prefer.

# Visualize

Fluidigmr does not provide plotting function, because all the functions that you need to plot this kind of data are already available in [ggplot2](https://ggplot2.tidyverse.org/). Indeed the dataset output of is already in a [tidy/gathered format](http://r4ds.had.co.nz/tidy-data.html).

Here we provide some ideas of how you can explore, plot and visualize your Fluidigm data.

## What variables are in your data?

Make the variables in your data explicit.

Effective visualization depends from the variable that determine the effect in your dataset and eventually by the question that you are trying to anwser.

For example in the sample dataset we are measuring the expression of four genes in 4 developmental stages in 5 species. 
The column `target_name` records the "gene" variable, while the other two variables, developmental stage and species, are hidden in the `sample_name` variable. We can make them explicit. 

The `sample_name` variable is encoded as "J-N1R1":

- The first letter "J" is the species.
- The letter and the number after the hyphen "N1" encode for the stage.
- The last letter and number "R1" encode for the replicate.

```{r}
fluidigm_expr <- 
  fluidigm_expr %>%
  # make species explicit
  mutate(species = str_split_fixed(string = sample_name,
                                   pattern = "-",
                                   n = 2)[, 1]) %>%
  # make stage explicit
  mutate(stage = str_sub(string = sample_name,
                         start = 3,
                         end = 4)) %>%
  # make replicate explicit
  mutate(replicate = str_sub(string = sample_name,
                         start = 5,
                         end = 6))
```

## Lineplot

A lineplot can be informative directly on normalized data, without scaling.

You can use facets to display multivariate data.

```{r, fig.height=6}
p <- fluidigm_expr %>%
  ggplot(aes(x = stage,
             y = expression)) +
  # plot each expression value into a point
  geom_point(size = 2,
               col = "#DA614D",
               alpha = .8) +
  # it might be complicated to add a line that connect means
  # across stages,
  # because stages are encoded as a categorical variable,
  # we must turn them into numeric
  stat_summary(aes(x = stage %>%
                     as.factor(.) %>%
                     as.numeric(.)),
               fun.y = mean,
               geom="line",
               size = 1.2,
               alpha = .7,
               linejoin = "round") +
  # implement facetting with independent y values
  # for each genes, since expression values are
  # not comparable across genes
  facet_grid(target_name ~ species,
               scales = "free_y") +
  theme_bw() 

p %>% print()
```

Some patterns in your data might be more evident if you plot them in log  scale.

```{r, warning=FALSE, fig.height=6}
p_log <- 
  p + scale_y_log10()

p_log %>% print()
```

Also using heatmap with ggplot2

```{r, fig.height=2}
p_heat <- 
  fluidigm_expr %>%
  ggplot(aes(x = stage,
             y = target_name,
             fill = expression)) +
  geom_tile() +
  facet_grid(. ~ species) +
  scale_fill_viridis_c()

p_heat %>% print()
```

Sometime heatmap don't convey muich invormation if you don't scale your data.

For example we can scale every gene on z-score to make them comparable.

```{r}
scaled_gene <- 
  fluidigm_expr %>% 
  group_by(target_name) %>%
  mutate(scaled_expression = expression %>%
           rescale(from = range(.), to = c(0,1)))
```

Check if the heatmap has improved

**remeber to map the colours to the new `scaled_expression` values**

```{r, fig.height=2}
p_heat <- 
  scaled_gene %>%
  ggplot(aes(x = stage,
             y = species,
             fill = scaled_expression)) +
  geom_tile() +
  facet_grid(. ~ target_name) +
  scale_fill_viridis_c()

p_heat %>% print()
```

We can try to set more groups.

```{r}
scaled_dat <- 
  fluidigm_expr %>%
  group_by(target_name, species) %>%
  mutate(scaled_expression = expression %>%
           rescale(from = range(.), to = c(0,1)))
```

Check if the heatmap has improved

**remeber to map the colours to the new `scaled_expression` values**

```{r,  fig.height=2}
p_heat <- 
  scaled_dat %>%
  ggplot(aes(x = stage,
             y = species,
             fill = scaled_expression)) +
  geom_tile() +
  facet_grid(. ~ target_name) +
  scale_fill_viridis_c()

p_heat %>% print()
```

# Scale and center on a specific group - Timecourses

Sometime it might be useful to center the expression data on a specific timepoint or condition, in a manner similar to the $2^{ - \Delta \Delta Ct}$ method

This can be achieved with the function `center_stage()`:

```{r}
center_stage <- function(scaled_data, group_var, group_var2, center_var)
  { 
  stopifnot("scaled_expression" %in% colnames(scaled_data))
  group_var <- enquo(group_var)
  group_var2 <- enquo(group_var2)
    scaled_data %>% 
      group_by(!!group_var, !!group_var2) %>% 
      mutate(center_mean = case_when(stage == center_var ~ scaled_expression) %>%
               mean(na.rm = T),
             centered_exp = scaled_expression/center_mean) 
}

centered_dat <- 
  scaled_dat %>% 
  center_stage(group_var = species,
               group_var2 = target_name,
               center_var = "N1")

centered_dat %>% 
  ggplot(aes(x = stage,
             y = centered_exp,
             group = species,
             colour = species)) +
  # plot each expression value into a point
  # geom_point(size = 2,
  #              col = "#DA614D",
  #              alpha = .8) +
  # it might be complicated to add a line that connect means
  # across stages,
  # because stages are encoded as a categorical variable,
  # we must turn them into numeric
  stat_summary(aes(x = stage %>%
                     as.factor(.) %>%
                     as.numeric(.)),
               fun.y = mean,
               geom="line",
               size = 1.2,
               alpha = .7,
               linejoin = "round") +
  # implement facetting with independent y values
  # for each genes, since expression values are
  # not comparable across genes
  facet_wrap(facet = "target_name",
             scales = "free",
             nrow = 1) +
  theme_bw() 
```

